---
title: "lab6"
format: pdf
author: "Angelica Rock (PID 15781397)"
---

# My first function :)

```{r}
add <- function(x,y=1){
  x+y
}
```

# Can I just use it?

```{r}
add(1,1)
add(1,100)
add(c(100,1,100), 1)
add(10, 10)
add(10)
```

# Q: Make a function "Generate_DNA()" that makes a random nucleotide sequence of any length (user specifies the length)

```{r}
  bases <- c("A", "C", "G", "T")
  sequence <- sample(bases, size=100, replace = TRUE)
```

Above is my working snippet, now I can make it into a function

```{r}
Generate_DNA <- function(length) {
  bases <- c("A", "C", "G", "T")
  sequence <- sample(bases, size=length, replace = TRUE)
  return(sequence)
}
```

Now I can run this function to get output

```{r}
Generate_DNA(10)
```

# the paste function is included here because collapse = "" changes the output from "A" "B" "C" etc. to ABC to put the output together so you can copy and paste it into BLAST to get the identities

```{r}
Generate_protein <- function(length) {
  amino_acids <- unique(bio3d::aa.table$aa1)[1:20]
  sequence <- sample(amino_acids, size=length, replace = TRUE)
  sequence <- paste(sequence, collapse ="")
  return(sequence)
}
```

Run function

```{r}
Generate_protein(10)
```

I want to generate sequences of length 6 to 12

```{r}
answer <- sapply(6:12, Generate_protein)
```

```{r}
cat(paste(">id.", 6:12, "\n", sep = "", answer), sep="\n")
```
